Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
report fewer digits on % change estimates
(This seemed like false precision)

The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
In 2010, 30-day-mean meteorological anomalies between February and May included lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD)<!--, including in the wet-season [KAT: commenting this out because I find it unclear/ don't understand what it means]-->.
<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet-season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
In 2020, meteorological anomalies were pronounced both in the dry- and wet-seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry- and wet-seasons.

@@

<!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
Drought years were associated with lower plot-level growth and stronger tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. 
<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. -->
<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)` respectively. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

>>>>

The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
In 2010, 30-day-mean meteorological anomalies between February and May included lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD)<!--, including in the wet-season [KAT: commenting this out because I find it unclear/ don't understand what it means]-->.
<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet-season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
1- to 12-month SPEI fell below -2, lowest in the dendroband census period and indicative of severe drought. 
In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
3- and 6-month SPEI values approached -2 in February, indicative of borderline severe drought.
12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry- and wet-seasons.

@@

<!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
Drought years were associated with lower plot-level growth and stronger tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. -->
<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)` respectively. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
